---
title: "class09"
author: "Niam Shasha A17805610"
format: pdf
toc: TRUE
---

## Data Import

```{r}
candy <- read.csv("candy-data.txt", row.names=1)
head(candy)
```
>Q1. How many different candy types are in this dataset?


```{r}
dim(candy)
```

there are 85 different candy types 


>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types

```{r}
candy["Twix", ]$winpercent
```


>Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way", ]$winpercent

```

The wninepercent value is 73.09956 of Milky Way. 

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent

```

The winepercent value is 76.7686

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?


```{r}

candy["Tootsie Roll Snack Bars", ]$winpercent

```


The winepercent value is 49.6535 of a tootsie roll.

```{r}
skimr::skim(candy)
```


```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

the variables and columns that are on different scales as compared to the majority of the other columns is winpercent this is because most vairables have a value of 0 or 1, and 0 to 1. However, winperecent ranges from values between 1 to 100 which is at a significantly higher scale than the other columns that are limited to values between 0 and 1. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

The zero represents that the candy has no chcolate and the one represents the candy has chocolate 

>Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20, fill="purple") +
  ggtitle("Distribution of Win Percent")
```


>Q9. Is the distribution of winpercent values symmetrical?

No, it seems to be slightly skewed with a slight right skew. 

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
median(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


The center of the distribution is approximately below 50% as the median is slightly below 50% at 47.83% the mean is at 50.32% which indicates that the center is approximaelty but more slightly below 50% as the median is below 50% and it represent the center of the data.

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?


```{r}
choc.ind <- as.logical(candy$chocolate)
choc.candy <- candy[choc.ind, ]
choc.win <-  choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.ind <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.ind,]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```
 
Chocolate candy is ranked higher than fruit candy at 60.92153 and fruity candy is ranked 44.11974. 

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

These are significantly significant because the two sample t-test p-values are less than 0.05 at 2.871e-08. 

>Q13. What are the five least liked candy types in this set?


```{r}
library(dplyr)

candy |>
  arrange(winpercent) |>
  select(winpercent) |>
  head(5)
```
 

```{r}
x <- c(5,10,1)
sort(x)
```

```{r}
head(candy[order(candy$winpercent), ])

```

 The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)

candy |>
  arrange(desc(winpercent)) |>
  select(winpercent) |>
  head(5)
```

The top 5 all time favorite candy are Reese's Peanut Butter cup, Reese's Miniatures,Twix, Kit Kat, Snickers.  

>Q15. Make a first barplot of candy ranking based on winpercent values.

Barplot and code of candy rankings are found below:

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

With the reorder() function:

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```



```{r, fig.height=15, fig.width=10}
mycols <- rep("black", nrow(candy))

#Chololate Candy
mycols[as.logical(candy$chocolate)] <- "chocolate"
 #fruity candy in pink
mycols[as.logical(candy$fruity)] <- "pink"
#candy bars un brown
mycols[as.logical(candy$bar)] <- "brown"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=mycols) +
  ylab("")
```

>Q17. What is the worst ranked chocolate candy?

sixlets are the worst ranked chocolate candy 

>Q18. What is the best ranked fruity candy?

The best ranked fruity candy is starburst 

# Taking a look at pricepercent

```{r}
# fruity candy in pink
mycols [as.logical(candy$fruity) ] <- "red" #"pink"
```


```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, 
      pricepercent, 
      label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, max.overlaps = )
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's minatures is the highest ranked in terms of winpercent for the least money. 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
 ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


The top 5 most expensive candy types are Nik L Nip, Nestle Smarties, Ring Pop, Hersheys Krackel, and Hersheys Milk Chocolate. The least popular of this set is Nik L Nip with a lower winpercent 22.44534. 

# Exploring the correleation structure

We can calculate the pair-wise correlation of all our columns 

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruit are the most anti-correleated. 


>Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

I predict that chocolate and fruity candy will feature heavily in the PC1 due to their strong opposing correlation. 

# PCA

In this case we want to be sure to set `scale-TRUE` because we have one variable `winpercent` that is on a very different scale than all others and would otherwise dominate our PCA results. 


```{r}
pca <- prcomp(candy, scale =TRUE)
summary (pca)
```

First major result figure is the "score plot" of PC1 vs PC2 - how diffferent candy are related to each other on our PC axis 

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=mycols) +
  geom_text_repel(size =3.3, col=mycols, max.overlaps=7)

```


The second major results figure from PCA is the so-called "loadings plot" 

```{r}
ggplot(pca$rotation)+
  aes(PC1, 
      reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```
 
>Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?


In the positive direction fruity, pluribus, and hard are vairabled strongly associated in the positive direction. This makes sense as it seems like PC1 seperated fruity candy to chocolate candy. This is also supported in the correleation plot as it shows how chocolate and fruity have a negative correleation in relation to each other. 


>Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A winning candy would be a chocolate based candy with features such as bar, peanuts,nougat, and caramel. As seen earlier choclate candies have higher winpercent values as compared to fruit candies. The correleation analysis showed how chocolate is positively associated with winpercent. PCA showed that chocolate and bar candies are opposite from fruit candies and PC1 showing how chocolate candies is associated with a higher popularity.









