---
title: Lab08
author: "Niam Shasha (A17805610)"
format: pdf
toc: true
---

##Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. 

Today we will analyze a biopsyy data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.

##DATA IMPORT

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names =1)
```

```{r}
head(wisc.df,3)
```


Make sure we remove or exclude the diagnosis column from the data-set that we use for further analysis

```{r}
wisc.data <-  wisc.df[,-1]
diagnosis <- as.factor(wisc.df$diagnosis)
```


>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```
569 observations

>Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```
212 are malignant 

>Q3. How many variables/features in the data are suffixed with `_mean`?

```{r}
colnames(wisc.data)
```

```{r}
length( grep("_mean", colnames(wisc.data), value=T))
```
10 are suffixed with mean. 

##Principal Component Analysis

We need to scale our data before PCA with the `scale=TRUE` ARGUMENT TO `PRCOMP()`. 

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```


```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```



>Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


```{r}
biplot(wisc.pr)
```


>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is very difficult to understand the PCA model as there is many overlap and clustering and it  is not clear on what the graph is depicting.


```{r}
library(ggplot2)

ggplot(wisc.pr$x)+
  aes(PC1,PC2, col=diagnosis)+
  geom_point()
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x)+
  aes(PC1,PC3, col=diagnosis)+
  geom_point()
```

I notice that the points are dependent on PC1 and that is what causes the division between malignant and benign. The seperation is on the PC1 axis. 

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```


```{r}
pve <- pr.var / sum(pr.var)
pve

plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?



##Hierachical Clustering

>Q10.  Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?


```{r}
wisc.hclust <- hclust( dist( scale(wisc.data) ) )
plot(wisc.hclust)
```



```{r}
head(wisc.pr$x)
```

##COmbining Methods

```{r}
d <- dist(wisc.pr$x[,1:4])
wisc.pr.hclust <- hclust(d, method="ward.D2")
plot(wisc.pr.hclust)
abline(h=70, col="red")
```


```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```

How does this clustering `grps` correspond to the expert `diagnosis`?

```{r}
table(diagnosis)
```

```{r}
table(diagnosis, grps)
```










