---
title: "Hwk class 12"
author: "Niam Shasha"
format: pdf
---

>Q13 Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.
 
 Hint: The read.table(),
summary() and boxplot() functions will likely be useful here. There is an example R script
online to be used ONLY if you are struggling in vein. Note that you can find the medium value
from saving the output of the boxplot() function to an R object and examining this object.
There is also the medium() and summary() function that you can use to check your
understanding.

```{r}
url <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"

data <- read.table(url, header = TRUE)
head(data)
```

```{r}
table(data$geno)
```


```{r}
summary(data$exp)
```


```{r}
tapply(data$exp, data$geno, median)
```

The file was read into R the sample size for each genotype is 108 for A/A, 233 for A/G, 121 for G/G. The median expression levels was  31.24847 for A/A, 25.06486 for A/G, and 20.07363 for G/G.

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.


```{r}
boxplot(data$exp ~ data$geno, main= "Expression by Genotype", xlab= "Genotype", ylab="Expression" )
```

The boxplot demonstrates that the A/A genotype has the highest expression value than A/G and G/G. This shows that the SNP does affect the expression of ORMDL3 as there are varying expression levels for each genotype. The G allele may be associated with lower expression based on the plot, more analysis would need to be conducted to fully conclude this statement. 





