---
title: "class07: Machine Learning 1"
author: "Niam Shasha A17805610"
format: pdf
toc: true
---

## Background

Today we will explore core machine learning methods thst are very popular in bioinformatics. These include **clustering** and **deminsionallity** reduction. 

## Kmeans clustering

The main function inn "base" R for k-means clustering called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To help us do this w can use the `rnorm()` function. 


```{r}
hist(rnorm(10000, mean=3))
```

```{r}
x <- c(rnorm(30, -3),rnorm(30,+3))
x
```

```{r}
z <- cbind(x=x, y=rev(x))
plot(z)
```

```{r}
p <- 1:5
rbind(p,p)
```

Now we can run `kmeans()` on this input `z` and see what the results look like. 

```{r}
km <- kmeans(z, centers = 2)
km
```

```{r}
attributes(km)
```

>Q. How many ppoints are in each cluster

30 points in each cluster

```{r}
km$size
```


>Q What componenet of your result object details cluster assignment/membership

```{r}
km$cluster
```


>Q What componenet of your result object details cluster center


```{r}
km$centers
```

>Q. Plot `z` colored by the kmeans cluster assighnment and add cluster centers as blue points

```{r}
plot(z, col=km$cluster)
points(km$centers, col="blue", pch=15)
```

>Q. Run a K-means clustering and plot the results asking for 4 clusters (k=4)?

```{r}
km4 <- kmeans(z, centers = 4)
plot(z, col=km4$cluster)
points(km4$centers, col= "black", pch = 15)
```

>**N.B** You need to tell K-means the number of clusters (`CENTERS=2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for(i in 1:10){
  km <- kmeans(z, centers=i)
  km$tot.withinss
  ans <- c(ans, km$tot.withinss)
}
ans
plot(ans, typ="o", 
     xlab="number of clusters",
     ylab= "Total Sum of Squares Distance")
```


## Hierarchical Clustering 

The main function is "base" R for hierarchical Clustering is called `hclust()`

This function does not take your "raw" data for clustering, You must first build a "distance matrix" from your data and pass this as input to `hclust()`


```{r}
d <- dist(z)
hc <- hclust(d)
hc
```



There is a bespoke `plot()` method ofr `hclust()` results object. 


```{r}
plot(hc)
abline(h=8, col="red")
```

```{r}
plot(z)
```



Once we have our `hclust` object (our "tree" of "cluster dendogram") we can *"cut"* the tree to reval the clustering pattern. 

```{r}
cutree(hc, h=8)
```


>Q. Make a plot of `z` with your hclust results(i.e colored by cluster membership)


```{r}
grps <- cutree(hc, k=2)
plot(z, col =grps)
```

## Hands on with PCA

PCA is a dimentionsallity reeducation method that is popular for revealing patterns in complex databases 

## Analysis of UK food
let's analyze data of the eating habits of folks from the UK to see if the patterns and trends that have some regions being distinct from others. 

## Data Import

Data is available in CSV format so we can use the `read.csv()` function

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
 
>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)

```
There are 17 rows and 5 colunns

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
dim(x)
```



```{r}
x <- read.csv(url, row.names = 1)
head(x)
```

```{r}
dim(x)
```


I prefer x <- read.csv(url, row.names = 1) because it imports the file right with the row names without having to manually reset it. The read.csv approach is more reliable when rerunning your code and more robust as the x <- x[,-1] also deletes columns. 

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))

```



```{r}
barplot(as.matrix(x), beside= F, col=rainbow(nrow(x)))

```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing beside= T to beside= F. 


>Q4. Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

```{r}
library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)

```

```{r}
head(x_long)
```


```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```
changing it from geom_col(position = "dodge") to geom_col(position = "stack") results in a stacked barplot figure. 

>Q5. Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

The code in this figure pairs() makes scatter plots comparing each country where each point represents a certain type of food. If it lies on the diagonal it infers that the two countries being compared have the same consumption in both countries. Points that are not centered on the diagonal line means a certain country consumes it more than the other one.

```{r}

library(pheatmap)

pheatmap( as.matrix(x) )
```

>Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

England, Wales, and Scotland cluster together meaning they have similar food consumption pattern. N Ireland clusters differently as they have a higher consumption of certain items such a fresh potatoes, however using the pairs and heatmap it is unfortunately not as easy to tell and I had to thus approximate what I see as visually different. 

>**Key-point**: Even relatively small datasets can prove challenging to interpret. 

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. The function wants the "observations" to be rows and the "variables" to be columns. 

```{r}
pca <- prcomp(t(x))
summary(pca)
```
The rturned `pca` object has components that we can use to make our main result figures. 

```{r}
attributes(pca)
```

The main result figure from this analysis is called a "PC score plot" or "ordenation plot" or "PC plot" or PC1 vs PC2 plot. 

The plot shows how samples relate to each other along our new acis.

This is our new "reducded-dimensional space". 


```{r}
pca$x
```

>Q7 . Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point()
```



```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()

```




## Tidy the data
fix anything that went wrong with data import

## Exporatory analysis

Make some plots to help make sense of ovious trends..

##PCA
```{r}
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x)+
  aes(PC1, PC2) +
  geom_point(col=mycols)
```

>  Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.



```{r}
pca$x
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x)+
  aes(PC1, PC2, label =row.names(pca$x)) +
  geom_point(col=mycols)+
geom_text(size=3, vjust=4, col=mycols)

```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v

```

```{r}
## or the second row here...
z <- summary(pca)
z$importance
```

```{r}
# Create scree plot with ggplot
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```



```{r}
pca$rotation
```


```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation), PC1) ) +
  geom_col()
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```


>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

The two food groups that feature prominently are soft drinks (positive) and fresh_potatoes (negative). PC2 shows the contrast between these food consumption (soft drinks and fresh poataoes)

